---
title: "Tower settlement context: exploring the measures"
---

```{r}
#| label: setup
#| include: false
source(here::here("scripts", "00_setup.R"))
library(cluster)   # silhouette widths; ships with R

FRAME_SNAPSHOT <- "dove_2026-09-19"   # <- named, never discovered

# ---- choices this chapter makes, all provisional ---------------------------
#
# None of these is a decision. They are starting points for the analytical
# team to calibrate, and they are set here so they are easy to find and vary.

# Bells at the two towers with two frame rings (Gresford, Rugby): Q-020, open.
MULTI_RING_BELLS <- "max"

# Measures the clustering sees. Logged population at four kernel scales, plus
# tower density at the local and wide scales. Bells are left out because they
# are close to towers x average ring size, and would double-count density.
CLUSTER_VARS <- c("pop_k1km", "pop_k2km", "pop_k4km", "pop_k8km",
                  "towers_k2km", "towers_k8km")

# Number of principal components fed to k-means, and the k shown in detail.
N_PCS      <- 3
K_CLUSTERS <- 6
K_RANGE    <- 3:10

# Rule-based alternative: bands of local (2 km kernel) and wider (8 km kernel)
# population. NULL means "use tertiles". Replace with round numbers once the
# distributions below have been looked at.
LOCAL_BREAKS <- NULL
WIDE_BREAKS  <- NULL

# Bell classes for the stratification cross-tab.
BELL_BREAKS <- c(4, 6, 8, 10, Inf)          # 4-5, 6-7, 8-9, 10+
BELL_LABELS <- c("4-5", "6-7", "8-9", "10+")

# Sample size for the expected-allocation table (chased towers, Lucky Dip).
N_CHASED <- 1000

set.seed(20261116)   # reference week; any fixed seed would do

# ---- palette ----------------------------------------------------------------
# Categorical: fixed slot order, never cycled (validated reference palette).
PAL_CAT <- c("#2a78d6", "#eb6834", "#1baf7a", "#eda100",
             "#e87ba4", "#008300", "#6250d6", "#e34948")
# Ordinal blue ramp for the Degree of Urbanisation, light (rural) to dark.
# Lightest step is 250, so the most rural class still clears 2:1 on white.
PAL_SMOD <- c("Very low density rural"   = "#86b6ef",
              "Low density rural"        = "#5598e7",
              "Rural cluster"            = "#2a78d6",
              "Suburban or peri-urban"   = "#256abf",
              "Semi-dense urban cluster" = "#1c5cab",
              "Dense urban cluster"      = "#184f95",
              "Urban centre"             = "#0d366b",
              "Water"                    = "#9a9a9a")

theme_set(
  theme_minimal(base_size = 11) +
    theme(panel.grid.minor = element_blank(),
          panel.grid.major = element_line(linewidth = 0.25, colour = "grey88"),
          legend.position = "bottom",
          plot.title.position = "plot")
)

lg  <- function(x) log10(x + 1)     # zeros exist (Lundy); +1 person is harmless
fmt <- scales::label_comma(accuracy = 1)
```

```{r}
#| label: load
#| include: false
cur <- \(stem) here::here("data", "curated", paste0(stem, "_", FRAME_SNAPSHOT, ".parquet"))

context   <- arrow::read_parquet(cur("tower_context"))
frame     <- arrow::read_parquet(cur("frame"))
locations <- arrow::read_parquet(cur("tower_locations"))
special   <- readr::read_csv(
  here::here("data", "reference", "special_towers_draft.csv"),
  col_types = readr::cols(.default = "c")
)

# Every measure the context table carries, kernels first, then radii.
measures <- grep("^pop_(k|r)\\d+km$", names(context), value = TRUE) |>
  sub(pattern = "^pop_", replacement = "")
measures <- measures[order(!startsWith(measures, "k"),
                           as.numeric(gsub("\\D", "", measures)))]

# One row per FRAME tower (D-017, D-019). Display and country from the first
# frame ring; bells per MULTI_RING_BELLS.
tw <- frame |>
  dplyr::filter(.data$frame_full) |>
  dplyr::summarise(
    bells   = if (MULTI_RING_BELLS == "max") max(.data$bells) else sum(.data$bells),
    country = dplyr::first(.data$country),
    display = dplyr::first(.data$display),
    .by = "tower_id"
  ) |>
  dplyr::inner_join(locations, by = "tower_id") |>
  dplyr::inner_join(context, by = "tower_id") |>
  dplyr::left_join(
    special |> dplyr::distinct(.data$tower_id, .keep_all = TRUE) |>
      dplyr::select("tower_id", special = "category"),
    by = "tower_id"
  ) |>
  dplyr::mutate(
    smod_class = factor(.data$smod_class, levels = names(PAL_SMOD)),
    bell_class = cut(.data$bells, BELL_BREAKS, BELL_LABELS, right = FALSE),
    region = dplyr::case_when(
      .data$country %in% c("England", "Wales") ~ "England & Wales",
      .data$country %in% BRITISH_ISLES ~ "Rest of British Isles",
      TRUE ~ .data$country
    )
  )

n_missing_context <- sum(frame$frame_full) - sum(frame$frame_full & frame$tower_id %in% context$tower_id)
```

This chapter looks at how the settlement measures from `scripts/08` vary and
covary across the `r fmt(nrow(tw))` frame towers on snapshot
`r FRAME_SNAPSHOT`. The aim is to choose a small set of classes or scales to
stratify the chased sample by, and to have something concrete for the
analytical team to calibrate, validate and name.

Nothing here is a decision. The choices at the top of the source are
starting points.

::: {.callout-note}
**What the measures are (D-026).** For each tower:

- `k<h>km` is an exponential kernel whose weight halves every *h* km. 2 km is
  the prior mode for a local catchment.
- `r<d>km` is a plain count within *d* km.
- `pop_` is resident population from GHSL, `towers_` and `bells_` count
  frame towers and bells (the tower itself included), and `pop_per_bell_`
  is the ratio.
- A Huff-type allocation that splits population between competing towers
  was considered and deferred (D-026). The ratios here measure competition
  at the tower, not at the resident.
:::

## Coverage

```{r}
#| label: coverage
tw |>
  dplyr::count(.data$region, name = "towers") |>
  dplyr::mutate(share = scales::percent(.data$towers / sum(.data$towers), 0.1)) |>
  knitr::kable()
```

`r n_missing_context` frame rings have no context row. This should be zero.
If it is not, script 08's scope or the coordinates need checking.

## How each measure is distributed

Each panel is one measure on a log scale. The kernels and radii should look
alike in shape; the question is how much they differ in spread.

```{r}
#| label: fig-dist
#| fig-height: 9
#| fig-cap: "Distribution of each measure across frame towers (log10 of value + 1)."
long <- tw |>
  dplyr::select("tower_id", dplyr::matches(paste0("^(pop|towers|pop_per_bell)_(",
                                                 paste(measures, collapse = "|"), ")$"))) |>
  tidyr::pivot_longer(-"tower_id", names_to = c("family", "measure"),
                      names_pattern = "^(pop_per_bell|pop|towers)_(.*)$") |>
  dplyr::mutate(measure = factor(.data$measure, levels = measures),
                family  = factor(.data$family, levels = c("pop", "towers", "pop_per_bell"),
                                 labels = c("Population", "Frame towers", "Population per bell")))

ggplot(long, aes(lg(value))) +
  geom_histogram(bins = 60, fill = PAL_CAT[1], colour = NA) +
  facet_grid(measure ~ family, scales = "free") +
  labs(x = "log10(value + 1)", y = "Towers") +
  theme(strip.text.y = element_text(angle = 0))
```

```{r}
#| label: tbl-quantiles
long |>
  dplyr::summarise(
    p05 = stats::quantile(.data$value, 0.05, na.rm = TRUE),
    p25 = stats::quantile(.data$value, 0.25, na.rm = TRUE),
    p50 = stats::quantile(.data$value, 0.50, na.rm = TRUE),
    p75 = stats::quantile(.data$value, 0.75, na.rm = TRUE),
    p95 = stats::quantile(.data$value, 0.95, na.rm = TRUE),
    .by = c("family", "measure")
  ) |>
  dplyr::arrange(.data$family, .data$measure) |>
  dplyr::mutate(dplyr::across(p05:p95, \(x) signif(x, 3))) |>
  knitr::kable(caption = "Quantiles across frame towers")
```

## Do the radii add anything to the kernels?

If a radius ranks towers almost exactly as a kernel does, the radius can be
used to explain the kernel to a meeting, and nothing is lost. The table gives,
for each kernel, the radius with the highest rank correlation.

Expect the matching radius to be much wider than the half-distance. Over
uniform ground, half of a kernel's total weight lies within about 2.4
half-distances, and 90% within about 5.6. So a "1 km" kernel draws most of its
weight from within about 5 km. To a meeting, describe the 2 km kernel as
"people within about 5 km, nearer ones counting more", not as "people within
2 km".

```{r}
#| label: tbl-kernel-radius
pop_mat <- tw |> dplyr::select(dplyr::all_of(paste0("pop_", measures)))
rho <- stats::cor(pop_mat, method = "spearman")
kern <- paste0("pop_", grep("^k", measures, value = TRUE))
rad  <- paste0("pop_", grep("^r", measures, value = TRUE))

tibble::tibble(kernel = kern) |>
  dplyr::mutate(
    best_radius = rad[apply(rho[kern, rad, drop = FALSE], 1, which.max)],
    spearman    = round(apply(rho[kern, rad, drop = FALSE], 1, max), 3)
  ) |>
  knitr::kable()
```

```{r}
#| label: fig-kernel-radius
#| fig-height: 4
#| fig-cap: "2 km kernel against the 2 km and 5 km radii. Points far off the band are towers where the radius edge cuts through a town."
tw |>
  dplyr::select("pop_k2km", "pop_r2km", "pop_r5km") |>
  tidyr::pivot_longer(c("pop_r2km", "pop_r5km"), names_to = "radius") |>
  ggplot(aes(lg(pop_k2km), lg(value))) +
  geom_point(size = 0.6, alpha = 0.25, colour = PAL_CAT[1]) +
  facet_wrap(~radius) +
  coord_cartesian(xlim = c(lg(stats::quantile(tw$pop_k2km, 0.002)), NA)) +
  labs(x = "log10 population, 2 km kernel", y = "log10 population, radius")
```

## Correlation structure

Rank correlations across every measure, clustered so that measures that move
together sit together. Large blocks of near-1 mean redundant measures; the
interesting structure is in the off-diagonal blocks.

```{r}
#| label: fig-corr
#| fig-height: 8
#| fig-cap: "Spearman correlations between all measures."
all_vars <- tw |>
  dplyr::select(dplyr::matches(paste0("^(pop|towers|bells|pop_per_bell)_(",
                                      paste(measures, collapse = "|"), ")$")))
cm <- stats::cor(all_vars, method = "spearman", use = "pairwise.complete.obs")
ord <- stats::hclust(stats::as.dist(1 - cm))$order
cm_long <- as.data.frame(as.table(cm)) |>
  dplyr::mutate(Var1 = factor(.data$Var1, levels = colnames(cm)[ord]),
                Var2 = factor(.data$Var2, levels = colnames(cm)[ord]))

ggplot(cm_long, aes(Var1, Var2, fill = Freq)) +
  geom_tile(colour = "white", linewidth = 0.3) +
  scale_fill_gradient2(low = "#eb6834", mid = "#f2f2f2", high = "#184f95",
                       midpoint = 0, limits = c(-1, 1), name = "Spearman") +
  coord_equal() +
  labs(x = NULL, y = NULL) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        panel.grid = element_blank())
```

## Towers with almost nobody around them

A handful of towers have close to zero population within the 2 km kernel.
They are listed here, then kept out of the axis ranges below so they do not
squash the plots. They stay in every table and in the clustering.

```{r}
#| label: tbl-empty
tw |>
  dplyr::filter(.data$pop_k2km < 50) |>
  dplyr::arrange(.data$pop_k2km) |>
  dplyr::transmute(.data$display, .data$country, pop_k2km = round(.data$pop_k2km),
                   pop_k8km = fmt(.data$pop_k8km), .data$special) |>
  knitr::kable(caption = "Frame towers with fewer than 50 people in the 2 km kernel")

# Axis floor for the scatter plots: the 0.2nd percentile of the 2 km kernel.
floor_2km <- lg(stats::quantile(tw$pop_k2km, 0.002))
floor_8km <- lg(stats::quantile(tw$pop_k8km, 0.002))
```

## Local scale against wider scale

This plane carries most of your three phenomena:

- **Up the diagonal:** hamlet, then village, then town, then city.
- **Above the diagonal:** low local population but high wider population. A
  village on the edge of a conurbation, or a tower in a merged-villages
  suburb.
- **Below the diagonal:** a free-standing town with empty country around it.

```{r}
#| label: fig-local-wide
#| fig-height: 6
#| fig-cap: "Local (2 km kernel) against wider (8 km kernel) population, coloured by GHSL Degree of Urbanisation."
ggplot(tw, aes(lg(pop_k2km), lg(pop_k8km), colour = smod_class)) +
  geom_point(size = 0.8, alpha = 0.6) +
  scale_colour_manual(values = PAL_SMOD, drop = TRUE, name = NULL) +
  guides(colour = guide_legend(override.aes = list(size = 3, alpha = 1), nrow = 2)) +
  coord_cartesian(xlim = c(floor_2km, NA), ylim = c(floor_8km, NA)) +
  labs(x = "log10 population, 2 km kernel", y = "log10 population, 8 km kernel")
```

The vertical distance from the main band is a candidate single index of
"embeddedness": how much bigger the wider area is than the place itself.

```{r}
#| label: fig-embedded
#| fig-height: 4
#| fig-cap: "Embeddedness, log10(8 km kernel) minus log10(2 km kernel), by Degree of Urbanisation."
tw |>
  dplyr::mutate(embedded = lg(.data$pop_k8km) - lg(.data$pop_k2km)) |>
  ggplot(aes(smod_class, embedded)) +
  geom_boxplot(outlier.size = 0.4, fill = "#cde2fb", colour = "#184f95") +
  coord_flip() +
  labs(x = NULL, y = "log10(8 km) - log10(2 km)")
```

## Competition: people per tower and per bell

City centres should stand out as many towers and few residents. Dense
conurbations should show more towers than their population alone predicts.

```{r}
#| label: fig-competition
#| fig-height: 6
#| fig-cap: "Local population against local tower density (2 km kernel). Diagonal lines are constant population per tower."
ref <- tibble::tibble(ppt = c(100, 1000, 10000, 1e5))
ggplot(tw, aes(lg(towers_k2km), lg(pop_k2km))) +
  geom_abline(data = ref, aes(intercept = log10(ppt), slope = 1),
              colour = "grey70", linewidth = 0.3, linetype = 2) +
  geom_point(aes(colour = smod_class), size = 0.8, alpha = 0.6) +
  geom_point(data = dplyr::filter(tw, !is.na(special)),
             shape = 21, size = 2.2, stroke = 0.6, colour = "black", fill = NA) +
  scale_colour_manual(values = PAL_SMOD, drop = TRUE, name = NULL) +
  guides(colour = guide_legend(override.aes = list(size = 3, alpha = 1), nrow = 2)) +
  coord_cartesian(ylim = c(floor_2km, NA)) +
  labs(x = "log10 frame towers, 2 km kernel", y = "log10 population, 2 km kernel",
       caption = "Black rings: towers on the draft special-case list. Dashed lines: 100, 1,000, 10,000, 100,000 people per tower.")
```

```{r}
#| label: fig-ppb-smod
#| fig-height: 4
#| fig-cap: "Population per bell (2 and 8 km kernels) by Degree of Urbanisation."
tw |>
  dplyr::select("smod_class", "pop_per_bell_k2km", "pop_per_bell_k8km") |>
  tidyr::pivot_longer(-"smod_class") |>
  ggplot(aes(smod_class, lg(value))) +
  geom_boxplot(outlier.size = 0.4, fill = "#cde2fb", colour = "#184f95") +
  facet_wrap(~name) +
  coord_flip() +
  labs(x = NULL, y = "log10 population per bell")
```

## The off-the-shelf classes

How the GHSL Degree of Urbanisation and the ONS 2021 Rural Urban
Classification split the frame, and how they line up with each other.

```{r}
#| label: tbl-smod-bells
tw |>
  dplyr::count(.data$smod_class, .data$bell_class) |>
  tidyr::pivot_wider(names_from = "bell_class", values_from = "n", values_fill = 0, names_sort = TRUE) |>
  knitr::kable(caption = "Frame towers by Degree of Urbanisation and bells")
```

```{r}
#| label: tbl-smod-ruc
tw |>
  dplyr::filter(!is.na(.data$ruc21nm)) |>
  dplyr::count(.data$ruc21nm, .data$smod_class) |>
  tidyr::pivot_wider(names_from = "smod_class", values_from = "n", values_fill = 0, names_sort = TRUE) |>
  knitr::kable(caption = "England & Wales: ONS RUC 2021 against Degree of Urbanisation")
```

## Principal components

Principal components on the standardised logs of the clustering variables.
If the first component is "size" and the second is "local against wider",
the plane above already holds most of the information.

```{r}
#| label: pca
X <- tw |>
  dplyr::select(dplyr::all_of(CLUSTER_VARS)) |>
  dplyr::mutate(dplyr::across(dplyr::everything(), lg)) |>
  scale()
pca <- stats::prcomp(X)

tibble::tibble(
  component = paste0("PC", seq_along(pca$sdev)),
  variance  = scales::percent(pca$sdev^2 / sum(pca$sdev^2), 0.1),
  cumulative = scales::percent(cumsum(pca$sdev^2) / sum(pca$sdev^2), 0.1)
) |>
  knitr::kable(caption = "Variance explained")

round(pca$rotation[, 1:min(4, ncol(X))], 2) |>
  knitr::kable(caption = "Loadings")
```

```{r}
#| label: fig-pca
#| fig-height: 6
#| fig-cap: "Towers on the first two components, coloured by Degree of Urbanisation."
scores <- as.data.frame(pca$x[, 1:N_PCS])
ggplot(dplyr::bind_cols(tw, scores), aes(PC1, PC2, colour = smod_class)) +
  geom_point(size = 0.8, alpha = 0.6) +
  scale_colour_manual(values = PAL_SMOD, drop = TRUE, name = NULL) +
  guides(colour = guide_legend(override.aes = list(size = 3, alpha = 1), nrow = 2))
```

## Candidate classification A: clusters

k-means on the first `r N_PCS` components, for k from `r min(K_RANGE)` to
`r max(K_RANGE)`. The elbow and the mean silhouette width are guides, not
verdicts. What matters more is whether the clusters can be named and
explained.

```{r}
#| label: fig-k
#| fig-height: 3.5
#| fig-cap: "Within-cluster sum of squares and mean silhouette width by k (silhouette on a random 2,500 towers)."
km_fits <- purrr::map(K_RANGE, \(k) stats::kmeans(scores, centers = k, nstart = 50, iter.max = 100))
sil_idx <- sample(nrow(scores), min(2500, nrow(scores)))
d_sub   <- stats::dist(scores[sil_idx, ])

k_diag <- tibble::tibble(
  k = K_RANGE,
  within_ss = purrr::map_dbl(km_fits, "tot.withinss"),
  silhouette = purrr::map_dbl(km_fits, \(f)
    mean(cluster::silhouette(f$cluster[sil_idx], d_sub)[, "sil_width"]))
) |>
  tidyr::pivot_longer(-"k") |>
  dplyr::mutate(name = dplyr::recode(.data$name,
    within_ss = "Total within-cluster sum of squares",
    silhouette = "Mean silhouette width"))

ggplot(k_diag, aes(k, value)) +
  geom_line(colour = PAL_CAT[1], linewidth = 0.7) +
  geom_point(colour = PAL_CAT[1], size = 2) +
  facet_wrap(~name, scales = "free_y") +
  scale_x_continuous(breaks = K_RANGE) +
  labs(y = NULL)
```

```{r}
#| label: clusters
fit <- km_fits[[which(K_RANGE == K_CLUSTERS)]]

# Label clusters in a stable order (by median wider population, then local),
# so C1 is always the most rural. The labels are placeholders for the
# analytical team to name.
tw$cluster_raw <- fit$cluster
cl_order <- tw |>
  dplyr::summarise(w = stats::median(.data$pop_k8km), l = stats::median(.data$pop_k2km),
                   .by = "cluster_raw") |>
  dplyr::arrange(.data$w, .data$l) |>
  dplyr::mutate(cluster = factor(paste0("C", dplyr::row_number()),
                                 levels = paste0("C", seq_len(K_CLUSTERS))))
tw <- tw |>
  dplyr::left_join(cl_order |> dplyr::select("cluster_raw", "cluster"), by = "cluster_raw")
PAL_CL <- stats::setNames(PAL_CAT[seq_len(K_CLUSTERS)], levels(tw$cluster))
```

```{r}
#| label: tbl-cluster-profile
modal <- \(x) names(sort(table(x), decreasing = TRUE))[1]
tw |>
  dplyr::summarise(
    towers        = dplyr::n(),
    share         = scales::percent(dplyr::n() / nrow(tw), 0.1),
    pop_2km       = fmt(stats::median(.data$pop_k2km)),
    pop_8km       = fmt(stats::median(.data$pop_k8km)),
    towers_2km    = round(stats::median(.data$towers_k2km), 1),
    pop_per_bell_2km = fmt(stats::median(.data$pop_per_bell_k2km)),
    modal_smod    = modal(.data$smod_class),
    median_bells  = stats::median(.data$bells),
    .by = "cluster"
  ) |>
  dplyr::arrange(.data$cluster) |>
  knitr::kable(caption = "Cluster profiles (medians)")
```

Five towers drawn at random from each cluster, as a first aid to naming:

```{r}
#| label: tbl-cluster-examples
tw |>
  dplyr::slice_sample(n = 5, by = "cluster") |>
  dplyr::arrange(.data$cluster) |>
  dplyr::summarise(examples = paste(.data$display, collapse = "; "), .by = "cluster") |>
  knitr::kable()
```

```{r}
#| label: fig-cluster-plane
#| fig-height: 6
#| fig-cap: "Clusters on the local-against-wider plane."
ggplot(tw, aes(lg(pop_k2km), lg(pop_k8km), colour = cluster)) +
  geom_point(size = 0.8, alpha = 0.6) +
  scale_colour_manual(values = PAL_CL, name = NULL) +
  guides(colour = guide_legend(override.aes = list(size = 3, alpha = 1), nrow = 1)) +
  coord_cartesian(xlim = c(floor_2km, NA), ylim = c(floor_8km, NA)) +
  labs(x = "log10 population, 2 km kernel", y = "log10 population, 8 km kernel")
```

```{r}
#| label: fig-cluster-map
#| fig-height: 9
#| fig-cap: "Clusters mapped: Great Britain, with London, the West Midlands and Norfolk enlarged."
gb <- dplyr::filter(tw, .data$country %in% c("England", "Wales", "Scotland"))
ggplot(gb, aes(lon, lat, colour = cluster)) +
  geom_point(size = 0.5, alpha = 0.8) +
  scale_colour_manual(values = PAL_CL, name = NULL) +
  coord_quickmap() +
  guides(colour = guide_legend(override.aes = list(size = 3), nrow = 1)) +
  labs(x = NULL, y = NULL) +
  theme(axis.text = element_blank(), panel.grid = element_blank())
```

```{r}
#| label: fig-cluster-zoom
#| fig-height: 5
zooms <- tibble::tribble(
  ~area,            ~lon_min, ~lon_max, ~lat_min, ~lat_max,
  "London",           -0.55,     0.35,    51.25,    51.75,
  "West Midlands",    -2.35,    -1.55,    52.30,    52.70,
  "Norfolk",           0.40,     1.80,    52.35,    53.00
)
zoomed <- purrr::pmap(zooms, \(area, lon_min, lon_max, lat_min, lat_max)
  tw |>
    dplyr::filter(dplyr::between(.data$lon, lon_min, lon_max),
                  dplyr::between(.data$lat, lat_min, lat_max)) |>
    dplyr::mutate(area = area)) |>
  purrr::list_rbind()

ggplot(zoomed, aes(lon, lat, colour = cluster)) +
  geom_point(size = 1.4, alpha = 0.9) +
  scale_colour_manual(values = PAL_CL, name = NULL) +
  facet_wrap(~area, scales = "free") +
  guides(colour = guide_legend(override.aes = list(size = 3), nrow = 1)) +
  labs(x = NULL, y = NULL) +
  theme(axis.text = element_blank(), panel.grid = element_blank())
```

## Candidate classification B: a simple two-scale grid

Easier to explain than clusters. Bands of local population crossed with bands
of wider population, giving at most nine cells. With no breaks set, these are
tertiles.

```{r}
#| label: grid
band <- function(x, breaks, labels) {
  if (is.null(breaks)) breaks <- stats::quantile(x, c(1/3, 2/3))
  cut(x, c(-Inf, breaks, Inf), labels = labels, right = FALSE)
}
tw <- tw |>
  dplyr::mutate(
    local = band(.data$pop_k2km, LOCAL_BREAKS, c("small place", "mid", "large place")),
    wide  = band(.data$pop_k8km, WIDE_BREAKS,  c("sparse area", "mid", "dense area")),
    grid  = interaction(.data$local, .data$wide, sep = " / ", lex.order = TRUE)
  )

used_local <- if (is.null(LOCAL_BREAKS)) stats::quantile(tw$pop_k2km, c(1/3, 2/3)) else LOCAL_BREAKS
used_wide  <- if (is.null(WIDE_BREAKS))  stats::quantile(tw$pop_k8km, c(1/3, 2/3)) else WIDE_BREAKS

tw |>
  dplyr::count(.data$local, .data$wide) |>
  tidyr::pivot_wider(names_from = "wide", values_from = "n", values_fill = 0, names_sort = TRUE) |>
  knitr::kable(caption = "Frame towers in each grid cell (rows: local; columns: wider)")
```

Breaks used: local `r paste(fmt(used_local), collapse = " and ")` people
(2 km kernel); wider `r paste(fmt(used_wide), collapse = " and ")` people
(8 km kernel).

How the grid lines up with the clusters and with the Degree of Urbanisation:

```{r}
#| label: tbl-grid-compare
tw |>
  dplyr::count(.data$grid, .data$cluster) |>
  tidyr::pivot_wider(names_from = "cluster", values_from = "n", values_fill = 0, names_sort = TRUE) |>
  knitr::kable(caption = "Grid cells against clusters")

tw |>
  dplyr::count(.data$grid, .data$smod_class) |>
  tidyr::pivot_wider(names_from = "smod_class", values_from = "n", values_fill = 0, names_sort = TRUE) |>
  knitr::kable(caption = "Grid cells against Degree of Urbanisation")
```

## What each candidate means for stratification

Strata are candidate class x bell class. The table shows the frame count in
each cell and the expected number of chased towers under proportional
allocation at n = `r fmt(N_CHASED)`. Cells expecting fewer than about 10 give
unstable stratum estimates and would need collapsing.

```{r}
#| label: tbl-strata
strata_table <- function(class_col, label) {
  tw |>
    dplyr::count(class = .data[[class_col]], .data$bell_class) |>
    dplyr::mutate(expected = round(.data$n / nrow(tw) * N_CHASED, 1)) |>
    dplyr::summarise(
      candidate = label,
      cells = dplyr::n(),
      cells_under_10 = sum(.data$expected < 10),
      smallest_expected = min(.data$expected)
    )
}
dplyr::bind_rows(
  strata_table("smod_class", "Degree of Urbanisation (GHSL)"),
  strata_table("cluster",    paste0("Clusters, k = ", K_CLUSTERS)),
  strata_table("grid",       "Two-scale grid")
) |>
  knitr::kable(caption = "Stratum cell sizes, candidate x bell class")
```

```{r}
#| label: tbl-strata-detail
tw |>
  dplyr::count(.data$cluster, .data$bell_class) |>
  dplyr::mutate(cell = paste0(.data$n, " (", round(.data$n / nrow(tw) * N_CHASED), ")")) |>
  dplyr::select(-"n") |>
  tidyr::pivot_wider(names_from = "bell_class", values_from = "cell", values_fill = "0", names_sort = TRUE) |>
  knitr::kable(caption = "Clusters x bells: frame towers (expected chased towers)")
```

## Where the special-case towers fall

```{r}
#| label: tbl-special
tw |>
  dplyr::filter(!is.na(.data$special)) |>
  dplyr::count(.data$special, .data$cluster) |>
  tidyr::pivot_wider(names_from = "cluster", values_from = "n", values_fill = 0, names_sort = TRUE) |>
  knitr::kable(caption = "Draft special-case list against clusters")
```

If the City of London towers do not form, or sit inside, a distinct cluster,
that supports handling them by list rather than by measure.

## Outside the British Isles

The measures use the same definitions everywhere. Whether the clusters still
mean the same thing abroad is a separate question.

```{r}
#| label: tbl-abroad
tw |>
  dplyr::count(.data$region, .data$cluster) |>
  tidyr::pivot_wider(names_from = "cluster", values_from = "n", values_fill = 0, names_sort = TRUE) |>
  knitr::kable(caption = "Clusters by region")
```

## For the analytical team

Questions this chapter is meant to help answer:

1. **Kernels or radii?** If the radii track the kernels closely, use radii
   in anything said to a meeting.
2. **How many dimensions?** If two components carry most of the variance, a
   two-scale grid may communicate as well as clusters do.
3. **Which k, and what are the clusters called?** Use the example towers and
   the maps. Names should be things a ringer would recognise.
4. **Do the special-case towers need a list?** Or do the measures find them?
5. **Which candidate gives workable strata** once crossed with bells, at
   about 1,000 chased towers?
